---
title: Was the Husker Women's Basketball Team Worthy of Earning a Spot in the Top
  25?
author: Abby Post
date: '2022-04-08'
slug: was-the-husker-women-s-basketball-team-worthy-of-earning-a-spot-in-the-top-25
categories: []
tags: []
---
```{r message=FALSE, warning=FALSE, echo=FALSE }

library(wehoop)
library(tidyverse)
library(scales)
library(ggrepel)
library(cowplot)
library(gt)
```
```{r message=FALSE, warning=FALSE, echo=FALSE }
standings <- espn_wbb_standings(2022)
```
The Nebraska women’s basketball team had a season unlike any of its recent years. Many deemed it "special" with the success they had. Starting the year off with a 12-0 record, Nebraska had built up its confidence heading into Big Ten play. Led by Amy Williams the Huskers completed their season 24-9 with a trip to the NCAA tournament as an 8th seed. One thing the Husker team couldn’t mark off of its list of accomplishments was making it into the AP Top 25 Poll. Husker fans questioned the reasoning behind this all season long. Was this on purpose or was Nebraska overlooked?

Without looking at NCAA tournament wins and losses, Nebraska had a 24-8 record. When comparing its record to other teams that did earn spots in the top rankings, Nebraska was sixteenth in winning percentage. The Huskers came about two percent behind Iowa, who it faced and lost to three times during the season. Nebraska had a higher winning percentage compared to Oklahoma, Tennessee, and Arizona, who were ranked No. 18, No. 19, and No. 22, respectively, in the AP Poll at the end of the season.


```{r message=FALSE, warning=FALSE, echo=FALSE}
  wbb_box <- wehoop::load_wbb_team_box(2022) 
```
```{r message=FALSE, warning=FALSE, echo=FALSE}
standings <- standings %>% 
  mutate_at(3:13, as.numeric) %>% 
  mutate_at(40:41, as.numeric) %>% 
  mutate_at(14, as.numeric) %>% 
  select(team_id, team, wins, losses, winpercent, vsaprankedteams_wins, vsaprankedteams_losses, total)
```
```{r message=FALSE, warning=FALSE, echo=FALSE}
wbb <- wbb_box %>% 
  separate(free_throws_made_free_throws_attempted, into=c("FTmade", "FTattempted"), sep="-") %>%
  separate(field_goals_made_field_goals_attempted, into=c("FGmade", "FGattempted"), sep="-") %>%
  separate(three_point_field_goals_made_three_point_field_goals_attempted, into=c("3PTmade", "3PTattempted"), sep="-") %>%
  mutate_at(12:35, as.numeric) %>%
  mutate(
    team_score = ((FGmade-`3PTmade`) * 2) + (`3PTmade`*3) + FTmade,
    possessions = FGattempted - offensive_rebounds + turnovers + (.475 * FTattempted),
    team_offensive_efficiency = (team_score/possessions)*100
  )
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
teams <- c("South Carolina", "Louisville", "Stanford", "NC State", "UConn", "Texas", "Baylor", "Iowa", "LSU", "Iowa State", "Indiana", "Michigan", "Maryland", "Ohio State", "Kentucky", "Virginia Tech", "North Carolina", "Tennessee", "Arizona", "BYU", "Notre Dame", "Oklahoma", "Florida Gulf Coast", "UCF", "Princeton", "Nebraska")
  
```
```{r message=FALSE, warning=FALSE, echo=FALSE}
top25 <-wbb %>% inner_join(standings) %>% filter(team_short_display_name %in% teams) 

negames <- top25 %>% filter(team_short_display_name == "Nebraska")
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
top25 %>%
filter(team_short_display_name %in% teams) %>%
select(team_short_display_name, wins, losses, winpercent) %>%
  group_by(team_short_display_name) %>%
  summarize(
    wins = mean(wins),
    losses = mean(losses),
    winpercent = mean(winpercent)
  ) %>%
  top_n(20, wt=winpercent) %>%
  arrange(desc(winpercent)) %>%
  gt() %>%
  cols_label(
    team_short_display_name = "Team",
    wins = "Wins",
    losses = "Losses",
    winpercent = "Win Percentage"
  ) %>%
  tab_header(
    "title" = "Nebraska Stands with Top Teams",
    "subtitle" = "The Huskers season win percentage places among top 20 best teams this season."
  ) %>%
tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("title")
  ) %>% tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("subtitle")
  
  ) %>% tab_style(
     locations = cells_column_labels(columns = everything()),
     style = list(
       cell_borders(sides = "bottom", weight = px(3)),
       cell_text(weight = "bold", size=12)
     )
   ) %>%
tab_source_note(
  source_note = md("By: Abby Post  | Source: wehoop: The SportsDataverse's R Package for Women's Basketball Data")) %>%
   opt_row_striping() %>% 
  opt_table_lines("none") %>%
    fmt_percent(
    columns = c(winpercent),
    decimals = 1
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = "red"),
      cell_text(color = "white")
      ),
    locations = cells_body(
      rows = team_short_display_name == "Nebraska")
  )
```
```{r message=FALSE, warning=FALSE, echo=FALSE}
notne <-top25 %>%
  group_by(team_short_display_name, team_id) %>%
  mutate()
  
ne <- negames %>%
  group_by(team_short_display_name, team_id)

```
```{r message=FALSE, warning=FALSE, echo=FALSE}
top25sb <-top25 %>% 
  group_by(team_short_display_name) %>%
  summarize(
    totalrankedwins = mean(vsaprankedteams_wins),
    totalrankedlosses = mean(vsaprankedteams_losses))
    
```
```{r message=FALSE, warning=FALSE, echo=FALSE}
top25sblonger<-top25sb %>% 
  pivot_longer(
    cols=starts_with("total"), 
    names_to="Type", 
    values_to="count")

neblonger <- top25sblonger %>% filter(team_short_display_name == "Nebraska")

```
Some of Nebraska's other successes that some may say have proved it worthy of a spot in the AP Poll during the season came from wins over ranked opponents. The first big win was taking down at the time No. 8 ranked Michigan. Next, Nebraska defeated No. 5 Indiana in front of the its home crowd. Nebraska then also beat Michigan a second time while knocking them out of the Big Ten Tournament. During this game Michigan had snagged the No. 10 spot in the AP Poll. These three huge wins may have given reason for Nebraska to have been looked at in earning a spot in the rankings. Even if they were looked at, these wins weren't deemed enough to snag them a spot. 


```{r message=FALSE, warning=FALSE, echo=FALSE}
ggplot() + 
  geom_bar(data=top25sblonger, aes(x=reorder(team_short_display_name, count), weight=count, fill=Type)) +
  geom_bar(data=neblonger, aes(x=reorder(team_short_display_name, count), weight=count, fill=Type), color="red") +
  scale_fill_manual(values=c("black", "blue"), name="Type", labels=c("Ranked losses", "Ranked wins")) + 
  scale_y_continuous(breaks=c(2, 4, 6, 8, 10, 12)) +
  coord_flip() +
  labs(
    x = "Teams",
    y = "Games against ranked opponent",
    title="Nebraska Could Match up Well with Ranked Teams" , 
    subtitle="Nebraska beat more ranked teams than four other teams that landed\na spot in the Top 25 at the end of the season.", 
    caption="By: Abby Post  | Source: wehoop: The SportsDataverse's R Package\nfor Women's Basketball Data") +
  theme_minimal() +
     theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    plot.title.position = "plot")
```

When diving into actual play and not just the overall results of the season, Nebraska still can prove itself. When calculating offensive ratings for teams, it shows how a team's offense compares to others. Summed up, the offensive rating is how many points are scored, or given for every 100 possessions played. When comparing Nebraska's offensive rating to the teams in the end of the season Top 25, we see that Nebraska fits right in. Offense plays a large part in winning games and so this gives an idea of how the Husker's play may have compared to teams that they didn't have the opportunity to play.

```{r message=FALSE, warning=FALSE, echo=FALSE}
wbb <- wbb_box %>% 
  separate(free_throws_made_free_throws_attempted, into=c("FTmade", "FTattempted"), sep="-") %>%
  separate(field_goals_made_field_goals_attempted, into=c("FGmade", "FGattempted"), sep="-") %>%
  separate(three_point_field_goals_made_three_point_field_goals_attempted, into=c("3PTmade", "3PTattempted"), sep="-") %>%
  mutate_at(12:35, as.numeric) %>%
  mutate(
    team_score = ((FGmade-`3PTmade`) * 2) + (`3PTmade`*3) + FTmade,
    possessions = FGattempted - offensive_rebounds + turnovers + (.475 * FTattempted),
    team_offensive_efficiency = (team_score/possessions)*100
)
```
```{r message=FALSE, warning=FALSE, echo=FALSE}
ggplot() + 
  geom_line(data=notne, aes(x=game_date, y=team_score, group=team_short_display_name), color="grey") +
  geom_line(data=ne, aes(x=game_date, y=team_score), color="red") +
  theme_minimal() +
  labs(
    x = "Date",
    y = "Offensive Rating",
    title="Husker's Offense Among the Best", 
    subtitle="Nebraska's offensive rating throughout the season is comparable to many\nof the teams that ended the season in the AP Poll.",
    caption="By: Abby Post  | Source: wehoop: The SportsDataverse's R Package for Women's Basketball Data") +
     theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    plot.title.position = "plot")
  
```
The Nebraska women's basketball team may not have earned a spot that most could say it was worthy of at some point in the season. With the strong start, good record, and big wins that gave Nebraska a season to remember, in the end a spot in the AP Top 25 Poll doesn't win or lose games for a team. The Husker team may have a chance to earn a spot in the rankings next year as all five starters are returning for what hopefully is another historic season.

The code is found [here](https://github.com/abbypost9/Notebook/blob/main/Project1.Rmd).